---
title: "An introduction to cutpointr"
author: "Christian Thiele"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cutpointr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

**cutpointr** is an R package for tidy calculation of "optimal" cutpoints. It 
supports several methods for calculating cutpoints and includes several 
metrics that can be maximized or minimized by selecting a cutpoint. Additionally,
**cutpointr** can automatically bootstrap the variability of the optimal 
cutpoints and return out-of-bag estimates of various metrics.


## Features

- Calculation of "optimal" cutpoints in binary classification tasks
- Tidy output, integrates well with functions from the tidyverse
- Bootstrapping for simulating the cutpoint variability and for getting 
out-of-bag estimates of various metrics (as a form of cross validation)
- Multiple methods for calculating cutpoints
- Multiple metrics can be chosen for maximization / minimization
- Standard/Nonstandard evaluation of the function arguments

## Calculating cutpoints

The included methods for calculating cutpoints are:

- `maximize_metric`: Maximize the metric function
- `minimize_metric`: Minimize the metric function
- `oc_manual`: Specify the cutoff value manually
- `oc_youden_kernel`: Maximize the Youden-Index after kernel smoothing
the distributions of the two classes
- `oc_youden_normal`: Maximize the Youden-Index parametrically
assuming normally distributed data in both classes
- `oc_OptimalCutpoints`: A tidy wrapper for optimal.cutpoints from the OptimalCutpoints package.
Supply an additional `oc_metric` argument with the method choice corresponding
to a method from the OptimalCutpoints package

The included metrics to be used with `minimize_metric` and `maximize_metric` are:

- `accuracy`: Fraction correctly classified
- `abs_d_sesp`: The absolute difference of sensitivity and specificity
- `kappa`: Cohen's Kappa
- `sum_sens_spec`: sensitivity + specificity
- `youden`: Youden- or J-Index = sensitivity + specificity - 1
- `odds_ratio`: (Diagnostic) odds ratio

**cutpointr** makes use of nonstandard evaluation for higher usability and to
allow for easy transformation of the variables. The inputs to the arguments
`method` and `metric` are functions so that user-defined functions can easily
be supplied instead of the built in ones.

## Applications

To showcase the functionality, we'll use the included `suicide` data set.

```{r}
library(cutpointr)
data(suicide)
head(suicide)
opt_cut <- cutpointr(suicide, dsi, suicide)
opt_cut
```

**cutpointr** makes assumptions about the direction of the dependency between 
class and x, if `direction` and / or `pos_class` or `neg_class` are not
specified. The same result can be achieved by manually defining `direction` and
the positive / negative classes which is slightly faster:

```{r}
opt_cut <- cutpointr(suicide, dsi, suicide, direction = ">=", pos_class = "yes",
                     neg_class = "no", method = maximize_metric, metric = youden)
opt_cut
```

`opt_cut` is a tidy data frame that returns the input data in a 
nested tibble. Methods for summarizing and plotting the data and
results are included:

```{r}
summary(opt_cut)
plot(opt_cut)
```

Predictions for new data can be made using `predict`:

```{r}
predict(opt_cut, newdata = data.frame(dsi = 0:5))
```

### Separate subgroups

Cutpoints can be separately estimated on a subgroup, gender in this case:

```{r}
opt_cut <- cutpointr(suicide, dsi, suicide, gender)
opt_cut
summary(opt_cut)
plot(opt_cut)
```

### Bootstrapping

If `boot_runs` is larger zero, **cutpointr** will carry out the usual cutpoint
calculation on the full sample, just as before, and additionally on 
`boot_runs` bootstrap samples. 

```{r}
set.seed(12)
opt_cut <- cutpointr(suicide, dsi, suicide, boot_runs = 200)
opt_cut
```

The returned object has the additional column `boot` which is a nested tibble that
includes the cutpoints per bootstrap sample along with the metric calculated using 
the function in `metric` and
a number of additional metrics. The values in the second column that are calculated
using the function in `metric` represent out-of-bag results. The other default
metrics are suffixed by `_b` to indicate in-bag results or `_oob` to indicate
out-of-bag results:

```{r}
opt_cut$boot
```

The summary and plots include additional elements:

```{r}
summary(opt_cut)
plot(opt_cut)
```

If a subgroup is given, the bootstrapping is carried out separately for every
subgroup:

```{r}
set.seed(12)
opt_cut <- cutpointr(suicide, dsi, suicide, gender, boot_runs = 200)
opt_cut
summary(opt_cut)
plot(opt_cut)
```

### Parallelized bootstrapping

Using `foreach` and `doRNG` the bootstrapping can easily be parallelized. The
`doRNG` package is being used to make the bootstrap sampling reproducible. It may
be preferable for long running tasks to specify `direction` and `pos_class` 
and / or `neg_class` manually to speed up `cutpointr`.

```{r}
if (suppressPackageStartupMessages(require(doSNOW) & require(doRNG))) {
  cl <- makeCluster(2) # 2 cores
  registerDoSNOW(cl)
  registerDoRNG(12) # Reproducible parallel loops using doRNG
  opt_cut <- cutpointr(suicide, dsi, suicide, gender, pos_class = "yes",
                 direction = ">=", boot_runs = 200, allowParallel = TRUE)
  stopCluster(cl)
  opt_cut
}
```

## Nonstandard evaluation

The arguments to `cutpointr` do not need to be enclosed in quotes. This is 
possible thanks to nonstandard evaluation of the arguments, which are 
evaluated in `data`. Alternatively, the arguments *can* be enclosed in quotes.
In that case, transforming the data within the function call is not possible
and `method` and `metric` functions that are enclosed in quotes are only looked
up within the **cutpointr** package. As an example of a transformation of the
`x`, `class` and `subgroup` variable consider:

```{r}
identical(cutpointr(suicide, dsi, suicide), cutpointr(suicide, "dsi", "suicide"))

set.seed(12)
opt_cut <- cutpointr(suicide, log(dsi + 1), suicide == "yes",
    subgroup = dsi %% 2 == 0, boot_runs = 30)
opt_cut
summary(opt_cut)
plot(opt_cut)
predict(opt_cut, newdata = data.frame(dsi = 0:5))
```

## cutpointr in the tidyverse

Since `cutpointr` outputs a tidy data frame and `data` is the first argument,
it can be conveniently used in conjunction with various functions from the "tidyverse". 

```{r}
opt_cut <- cutpointr(suicide, dsi, suicide)
class(opt_cut) # the result is also a data.frame

suppressMessages(library(dplyr))
suppressMessages(library(tidyr))
suppressMessages(library(purrr))
suppressMessages(library(ggplot2))
suicide %>%
    group_by(gender) %>%
    nest() %>%
    mutate(cutmod1 = map(data, function(x) cutpointr(x, dsi, suicide,
                                                     metric = accuracy)),
           cutmod2 = map(data, function(x) cutpointr(x, dsi, suicide,
                                                     metric = sum_sens_spec)))
```

## Plotting

**cutpointr** includes several convenience functions for plotting data from a 
`cutpointr` object. These include:

- `plot_cut_boot`: Plot the bootstrapped distribution of optimal cutpoints
- `plot_metric`: If `maximize_metric` or `minimize_metric` was used this function
plots all possible cutoffs on the x-axis vs. the respective metric values on
the y-axis
- `plot_metric_boot`: Plot the distribution of out-of-bag metric values
- `plot_precision_recall`: Plot the precision recall curve
- `plot_roc`: Plot the ROC curve
- `plot_x`: Plot the distribution of the predictor variable

```{r, fig.width=4, fig.height=3}
opt_cut <- cutpointr(suicide, dsi, suicide, gender, method = minimize_metric,
                     metric = abs_d_sesp, boot_runs = 100)
opt_cut
plot_cut_boot(opt_cut)
plot_metric(opt_cut)
plot_metric_boot(opt_cut)
plot_precision_recall(opt_cut)
plot_roc(opt_cut)
plot_x(opt_cut)
```


### Manual plotting

Since **cutpointr** returns a tidy data frame with the supplied data, bootstrap
results and the ROC curve in nested tibbles these data can coveniently be 
plotted manually. This offers additional ways of tweaking these plots as well
as the possibility to plot results that are not included in `plot`, 
`plot_metric`, `plot_cut_boot`, `plot_metric_boot` or `plot_roc`. The relevant
nested tibbles are in the columns `data`, `roc_curve` and `boot`. The following
is an example of accessing and plotting the grouped data.

```{r}
set.seed(123) # Some missing values expected
opt_cut <- cutpointr(suicide, dsi, suicide, gender, boot_runs = 100)
head(opt_cut$data)

opt_cut %>% 
    select(data, subgroup) %>% 
    unnest %>% 
    ggplot(aes(x = suicide, y = dsi)) + 
    geom_boxplot(alpha = 0.3) + facet_grid(~subgroup)
```

### ROC curve only

When running `cutpointr` a ROC curve is by default returned in the column `roc`.
This ROC curve can be plotted using `plot_roc`. Alternatively, if only the
ROC curve is desired and no cutpoint needs to be calculated, the ROC curve
can be plotted using the `roc` function and, for example, **ggplot2**. Note that
the `roc` function, unlike `cutpointr`, does not determine `direction`, `pos_class` or `neg_class`
automatically and does not support nonstandard evaluation, so the function
arguments have to be enclosed in quotation marks.

```{r}
roc(data = suicide, x = "dsi", class = "suicide", 
    pos_class = "yes", neg_class = "no") %>% 
    ggplot(aes(x = fpr, y = tpr)) + geom_line()
```


## User-defined method and metric functions

User defined functions can be supplied to method. 
To define a new method function, create a function that may take
as input(s):

- `data`: A data frame or tbl_df
- `x`: (character) The name of the predictor or independent variable
- `class`: (character) The name of the class or dependent variable
- `metric_func`: A function for calculating a metric, e.g. accuracy. Note
 that the method function does not necessarily have to accept this argument
- `pos_class`: The positive class
- `neg_class`: The negative class
- `direction`: ">=" if the positive class has higher x values, "<=" otherwise

The `...` argument can be used to avoid an error if not all of the above
arguments are needed and in order to pass additional arguments to `method`.
The function should return a data frame or tbl_df with
one row, the column "optimal_cutpoint", and an optinal column with an arbitraty name
with the metric value at the optimal cutpoint.

For example, a function for choosing the cutpoint as the mean of the independent
variable could look like this:

```{r, eval = FALSE}
mean_cut <- function(data, x, ...) {
    oc <- mean(unlist(data[, x]))
    return(data.frame(optimal_cutpoint = oc))
}
```

Since no metric is returned, `Sum_Sens_Spec`, the sum of sensitivity and 
specificity, is returned as the extra metric column in addition to accuracy, 
sensitivity and specificity.

Some `method` functions that make use of the additional arguments (that are 
captured by `...` in `mean_cut`) are already included in **cutpointr**, see
the list at the top. Since these functions are arguments to `cutpointr` 
their code can be accessed by simply typing their name, e.g.:

```{r}
oc_youden_normal
```


User defined **metric functions** can be used as well. They are mainly useful in
conjunction with `method = maximize_metric` or `method = minimize_metric`. 
In case of a different `method` function `metric` will only be used as the main
out-of-bag metric when plotting the result. The `metric` function should 
accept the following inputs as vectors:

- `tp`: Vector of true positives
- `fp`: Vector of false positives
- `tn`: Vector of true negatives
- `fn`: Vector of false negatives

The function should return a **matrix with one column**
and the inputs (`tp`, `fp`, `tn`, and `fn`) are **vectors**. If the column is named,
the name will be included in the output and plots. Avoid using names that
are identical to the column names that are by default returned by cutpointr.
The code of the included metric functions can be accessed by simply typing their name.

For example, this is the `accuracy` metric function:

```{r}
accuracy
```


