---
title: "An introduction to cutpointr"
author: "Christian Thiele"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{An introduction to cutpointr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(fig.width = 6, fig.height = 5, fig.align = "center")
options(rmarkdown.html_vignette.check_title = FALSE)
load("vignettedata/vignettedata.Rdata")
```

**cutpointr** is an R package for tidy calculation of "optimal" cutpoints. It 
supports several methods for calculating cutpoints and includes several 
metrics that can be maximized or minimized by selecting a cutpoint. Some of these
methods are designed to be more robust than the simple empirical optimization 
of a metric. Additionally,
**cutpointr** can automatically bootstrap the variability of the optimal 
cutpoints and return out-of-bag estimates of various performance metrics.

## Installation

You can install **cutpointr** from CRAN using the menu in RStudio or simply:

```{r CRAN, eval = FALSE}
install.packages("cutpointr")
```

## Example

For example, the optimal cutpoint for the included data set is 2 when maximizing the sum of sensitivity and specificity.

```{r}
library(cutpointr)
data(suicide)
head(suicide)
cp <- cutpointr(suicide, dsi, suicide, 
                method = maximize_metric, metric = sum_sens_spec)
```

```{r}
summary(cp)
```

```{r}
plot(cp)
```

When considering the optimality of a cutpoint, we can only make a judgement based
on the sample at hand. Thus, the estimated cutpoint may not be optimal within 
the population or on unseen data, which is why we sometimes put the "optimal" in
quotation marks.

`cutpointr` makes assumptions about the direction of the dependency between 
`class` and `x`, if `direction` and / or `pos_class` or `neg_class` are not
specified. The same result as above can be achieved by manually defining `direction` and
the positive / negative classes which is slightly faster, since the classes and direction
don't have to be determined:

```{r}
opt_cut <- cutpointr(suicide, dsi, suicide, direction = ">=", pos_class = "yes",
                     neg_class = "no", method = maximize_metric, metric = youden)
```

`opt_cut` is a data frame that returns the input data and the ROC curve
(and optionally the bootstrap results) in a 
nested tibble. Methods for summarizing and plotting the data and
results are included (e.g. `summary`, `plot`, `plot_roc`, `plot_metric`)

To inspect the optimization, the function of metric values per cutpoint can be
plotted using `plot_metric`, if an optimization function was used that returns
a metric column in the `roc_curve` column. For example, the `maximize_metric`
and `minimize_metric` functions do so:

```{r}
plot_metric(opt_cut)
```

Predictions for new data can be made using `predict`:

```{r}
predict(opt_cut, newdata = data.frame(dsi = 0:5))
```


## Features

- Calculation of optimal cutpoints in binary classification tasks
- Tidy output, integrates well with functions from the tidyverse
- Functions for plotting ROC curves, metric distributions and more
- Bootstrapping for simulating the cutpoint variability and for obtaining 
out-of-bag estimates of various metrics (as a form of internal validation)
with optional parallelisation
- Multiple methods for calculating cutpoints
- Multiple metrics can be chosen for maximization / minimization
- Tidyeval

# Calculating cutpoints

## Method functions for cutpoint estimation

The included methods for calculating cutpoints are:

- `maximize_metric`: Maximize the metric function
- `minimize_metric`: Minimize the metric function
- `maximize_loess_metric`: Maximize the metric function after LOESS smoothing
- `minimize_loess_metric`: Minimize the metric function after LOESS smoothing
- `maximize_spline_metric`: Maximize the metric function after spline smoothing
- `minimize_spline_metric`: Minimize the metric function after spline smoothing
- `maximize_gam_metric`: Maximize the metric function after smoothing via Generalized Additive Models
- `minimize_gam_metric`: Minimize the metric function after smoothing via Generalized Additive Models
- `maximize_boot_metric`: Bootstrap the optimal cutpoint when maximizing a metric
- `minimize_boot_metric`: Bootstrap the optimal cutpoint when minimizing a metric
- `oc_manual`: Specify the cutoff value manually
- `oc_mean`: Use the sample mean as the "optimal" cutpoint
- `oc_median`: Use the sample median as the "optimal" cutpoint
- `oc_youden_kernel`: Maximize the Youden-Index after kernel smoothing
the distributions of the two classes
- `oc_youden_normal`: Maximize the Youden-Index parametrically
assuming normally distributed data in both classes

## Metric functions

The included metrics to be used with the minimization and maximization methods 
are:

- `accuracy`: Fraction correctly classified
- `abs_d_sens_spec`: The absolute difference of sensitivity and specificity
- `abs_d_ppv_npv`: The absolute difference between positive predictive
value (PPV) and negative predictive value (NPV)
- `roc01`: Distance to the point (0,1) on ROC space
- `cohens_kappa`: Cohen's Kappa
- `sum_sens_spec`: sensitivity + specificity
- `sum_ppv_npv`: The sum of positive predictive value (PPV) and negative
predictive value (NPV)
- `prod_sens_spec`: sensitivity * specificity
- `prod_ppv_npv`: The product of positive predictive value (PPV) and 
negative predictive value (NPV)
- `youden`: Youden- or J-Index = sensitivity + specificity - 1
- `odds_ratio`: (Diagnostic) odds ratio
- `risk_ratio`: risk ratio (relative risk)
- `p_chisquared`: The p-value of a chi-squared test on the confusion
matrix
- `cost_misclassification`: The sum of the misclassification cost of
false positives and false negatives. Additional arguments: cost_fp, cost_fn
- `total_utility`: The total utility of true / false positives / negatives.
Additional arguments: utility_tp, utility_tn, cost_fp, cost_fn
- `F1_score`: The F1-score (2 * TP) / (2 * TP + FP + FN)
- `metric_constrain`: Maximize a selected metric given a minimal value of 
another selected metric
- `sens_constrain`: Maximize sensitivity given a minimal value of specificity
- `spec_constrain`: Maximize specificity given a minimal value of sensitivity
- `acc_constrain`: Maximize accuracy given a minimal value of sensitivity

Furthermore, the following functions are included which can be used as metric
functions but are more useful for plotting purposes, for example in 
`plot_cutpointr`, or for defining new metric functions: 
`tp`, `fp`, `tn`, `fn`, `tpr`, `fpr`, `tnr`, `fnr`, `false_omission_rate`,
`false_discovery_rate`, `ppv`, `npv`, `precision`, `recall`, `sensitivity`, and `specificity`.

The inputs to the arguments
`method` and `metric` are functions so that user-defined functions can easily
be supplied instead of the built-in ones.


## Separate subgroups and bootstrapping

Cutpoints can be separately estimated on subgroups that are defined by a third variable,
`gender` in this case. Additionally, 
if `boot_runs` is larger zero, `cutpointr` will carry out the usual cutpoint
calculation on the full sample, just as before, and additionally on 
`boot_runs` bootstrap samples. This offers a way of gauging the out-of-sample
performance of the cutpoint estimation method. If a subgroup is given, 
the bootstrapping is carried out separately for every
subgroup which is also reflected in the plots and output.

```{r separate subgroups and bootstrapping, eval = FALSE}
set.seed(12)
opt_cut_b <- cutpointr(suicide, dsi, suicide, boot_runs = 1000)
```
```{r}
opt_cut_b
```


The returned object has the additional column `boot` which is a nested tibble that
includes the cutpoints per bootstrap sample along with the metric calculated using 
the function in `metric` and
various default metrics. The 
metrics are suffixed by `_b` to indicate in-bag results or `_oob` to indicate
out-of-bag results:

```{r}
opt_cut_b$boot
```

The summary and plots include additional elements that summarize or display the
bootstrap results:

```{r}
summary(opt_cut_b)
plot(opt_cut_b)
```


### Parallelized bootstrapping

Using `foreach` and `doRNG` the bootstrapping can be parallelized easily. The
**doRNG** package is being used to make the bootstrap sampling reproducible.

```{r, eval = FALSE}
library(doParallel)
cl <- makeCluster(2) # 2 cores
registerDoParallel(cl)
registerDoRNG(12) # Reproducible parallel loops using doRNG
opt_cut <- cutpointr(suicide, dsi, suicide, gender, pos_class = "yes",
                     direction = ">=", boot_runs = 1000, allowParallel = TRUE)
stopCluster(cl)
opt_cut
```



# Additional features

## Finding all cutpoints with acceptable performance

By default, most packages only return the "best" cutpoint and disregard other cutpoints with quite similar performance, even if the performance differences are minuscule. **cutpointr** makes this process more explicit via the `tol_metric` argument. For example, if all cutpoints are of interest that achieve at least an accuracy within `0.05` of the optimally achievable accuracy, `tol_metric` can be set to `0.05` and also those cutpoints will be returned. 

In the case of the `suicide` data and when maximizing the sum of sensitivity and specificity, empirically the cutpoints 2 and 3 lead to quite similar performances. If `tol_metric` is set to `0.05`, both will be returned.

```{r}
library(tidyr)
library(dplyr)
opt_cut <- cutpointr(suicide, dsi, suicide, metric = sum_sens_spec, 
                     tol_metric = 0.05, break_ties = c)
opt_cut %>% 
    select(optimal_cutpoint, sum_sens_spec) %>% 
    unnest(cols = c(optimal_cutpoint, sum_sens_spec))
```


## Manual and mean / median cutpoints

Using the `oc_manual` function the optimal cutpoint will not be determined 
based on, for example, a metric but is instead set manually using the 
`cutpoint` argument. This is useful for supplying and evaluating cutpoints that were found
in the literature or in other external sources. 

The `oc_manual` function could also be used to set the cutpoint to the sample
mean using `cutpoint = mean(data$x)`. However, this may introduce bias into the
bootstrap validation procedure, since the actual mean of the population is
not known and thus the mean to be used as the cutpoint should be automatically determined in every resample.
To do so, the `oc_mean` and `oc_median` functions can be used.

```{r, eval = FALSE}
set.seed(100)
opt_cut_manual <- cutpointr(suicide, dsi, suicide, method = oc_manual, 
                       cutpoint = mean(suicide$dsi), boot_runs = 1000)
set.seed(100)
opt_cut_mean <- cutpointr(suicide, dsi, suicide, method = oc_mean, boot_runs = 1000)
```



## Nonstandard evaluation via tidyeval

The arguments to `cutpointr` do not need to be enclosed in quotes. This is 
possible thanks to nonstandard evaluation of the arguments, which are 
evaluated on `data`. 

Functions that use nonstandard evaluation are often not suitable for 
programming with. The use of nonstandard evaluation may lead to scoping 
problems and subsequent obvious as well as possibly subtle errors. 
**cutpointr** uses tidyeval internally and accordingly the same rules as 
for programming with `dplyr` apply. Arguments can be unquoted with `!!`:

```{r, eval = FALSE}
myvar <- "dsi"
cutpointr(suicide, !!myvar, suicide)
```



## Midpoints

So far - which is the default in `cutpointr` - we have considered all unique values of the predictor as possible cutpoints. An alternative could be to use a sequence of equidistant values instead, for example in the case of the `suicide` data all integers in $[0, 10]$. However, with very sparse data and small intervals between the candidate cutpoints (i.e. a 'dense' sequence like `seq(0, 10, by = 0.01)`) this leads to the uninformative evaluation of large ranges of cutpoints that all result in the same metric value. A more elegant alternative, not only for the case of sparse data, that is supported by **cutpointr** is the use of a mean value of the optimal cutpoint and the next highest (if `direction = ">="`) or the next lowest (if `direction = "<="`) predictor value in the data. The result is an optimal cutpoint that is equal to the cutpoint that would be obtained using an infinitely dense sequence of candidate cutpoints and is thus usually more efficient computationally. 

This behavior can be activated by setting `use_midpoints = TRUE`, which is the default. If we use this setting, we obtain an optimal cutpoint of 1.5 for the complete sample on the `suicide` data instead of 2 when maximizing the sum of sensitivity and specificity.

Assume the following small data set:

```{r}
dat <- data.frame(outcome = c("neg", "neg", "neg", "pos", "pos", "pos", "pos"),
                  pred    = c(1, 2, 3, 8, 11, 11, 12))
```

Since the distance of the optimal cutpoint (8) to the next lowest 
observation (3) is rather large we arrive at a range of possible cutpoints that
all maximize the metric. In the case of this kind of sparseness it might for example be
desirable to classify a new observation with a predictor value of 4 as belonging
to the negative class. If `use_midpoints` is set to `TRUE`, the mean of the 
optimal cutpoint and the next lowest observation is returned as the optimal
cutpoint, if direction is `>=`. The mean of the optimal cutpoint and the next
highest observation is returned as the optimal cutpoint, if `direction = "<="`.

```{r}
opt_cut <- cutpointr(dat, x = pred, class = outcome, use_midpoints = TRUE)
plot_x(opt_cut)
```

A simulation demonstrates more clearly that setting `use_midpoints = TRUE` avoids
biasing the cutpoints. To simulate the bias of the metric functions, the 
predictor values of both classes were drawn from normal distributions with 
constant standard deviations of 10, a constant mean of the negative class of 100
and higher mean values of the positive class that are selected in such a way 
that optimal Youden-Index values of 0.2, 0.4, 0.6, and 0.8 result in the population.
Samples of 9 different sizes were drawn and the cutpoints that maximize the 
Youden-Index were estimated. The simulation was repeated 10000 times. As can be
seen by the mean error, `use_midpoints = TRUE` eliminates the bias that is 
introduced by otherwise selecting the value of an observation as the optimal
cutpoint. If `direction = ">="`, as in this case, the observation that 
represents the optimal cutpoint is the highest possible cutpoint that leads to the
optimal metric value and thus the biases are positive. The methods `oc_youden_normal`
and `oc_youden_kernel` are always unbiased, as they don't select a cutpoint 
based on the ROC-curve or the function of metric values per cutpoint.

```{r, echo = FALSE}
plotdat_nomidpoints <- structure(list(sim_nr = c(1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 4L, 
4L, 4L, 4L, 4L, 4L, 4L, 4L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 6L, 
6L, 6L, 6L, 6L, 6L, 6L, 6L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 8L, 
8L, 8L, 8L, 8L, 8L, 8L, 8L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 10L, 
10L, 10L, 10L, 10L, 10L, 10L, 10L, 11L, 11L, 11L, 11L, 11L, 11L, 
11L, 11L, 12L, 12L, 12L, 12L, 12L, 12L, 12L, 12L, 13L, 13L, 13L, 
13L, 13L, 13L, 13L, 13L, 14L, 14L, 14L, 14L, 14L, 14L, 14L, 14L, 
15L, 15L, 15L, 15L, 15L, 15L, 15L, 15L, 16L, 16L, 16L, 16L, 16L, 
16L, 16L, 16L, 17L, 17L, 17L, 17L, 17L, 17L, 17L, 17L, 18L, 18L, 
18L, 18L, 18L, 18L, 18L, 18L, 19L, 19L, 19L, 19L, 19L, 19L, 19L, 
19L, 20L, 20L, 20L, 20L, 20L, 20L, 20L, 20L, 21L, 21L, 21L, 21L, 
21L, 21L, 21L, 21L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 23L, 
23L, 23L, 23L, 23L, 23L, 23L, 23L, 24L, 24L, 24L, 24L, 24L, 24L, 
24L, 24L, 25L, 25L, 25L, 25L, 25L, 25L, 25L, 25L, 26L, 26L, 26L, 
26L, 26L, 26L, 26L, 26L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 
28L, 28L, 28L, 28L, 28L, 28L, 28L, 28L, 29L, 29L, 29L, 29L, 29L, 
29L, 29L, 29L, 30L, 30L, 30L, 30L, 30L, 30L, 30L, 30L, 31L, 31L, 
31L, 31L, 31L, 31L, 31L, 31L, 32L, 32L, 32L, 32L, 32L, 32L, 32L, 
32L, 33L, 33L, 33L, 33L, 33L, 33L, 33L, 33L, 34L, 34L, 34L, 34L, 
34L, 34L, 34L, 34L, 35L, 35L, 35L, 35L, 35L, 35L, 35L, 35L, 36L, 
36L, 36L, 36L, 36L, 36L, 36L, 36L), method = structure(c(1L, 
2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 
2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 
2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 
2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 
2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 
2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 
2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 
2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 
2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 
2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 
2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 
2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 
2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 
2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 
2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 
2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 
2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 
2L, 3L, 4L, 5L, 6L, 7L, 8L, 1L, 2L, 3L, 4L, 5L, 6L, 7L, 8L), .Label = c("emp", 
"normal", "loess", "boot", "spline", "spline_20", "kernel", "gam"
), class = "factor"), n = c(30, 30, 30, 30, 30, 30, 30, 30, 50, 
50, 50, 50, 50, 50, 50, 50, 75, 75, 75, 75, 75, 75, 75, 75, 100, 
100, 100, 100, 100, 100, 100, 100, 150, 150, 150, 150, 150, 150, 
150, 150, 250, 250, 250, 250, 250, 250, 250, 250, 500, 500, 500, 
500, 500, 500, 500, 500, 750, 750, 750, 750, 750, 750, 750, 750, 
1000, 1000, 1000, 1000, 1000, 1000, 1000, 1000, 30, 30, 30, 30, 
30, 30, 30, 30, 50, 50, 50, 50, 50, 50, 50, 50, 75, 75, 75, 75, 
75, 75, 75, 75, 100, 100, 100, 100, 100, 100, 100, 100, 150, 
150, 150, 150, 150, 150, 150, 150, 250, 250, 250, 250, 250, 250, 
250, 250, 500, 500, 500, 500, 500, 500, 500, 500, 750, 750, 750, 
750, 750, 750, 750, 750, 1000, 1000, 1000, 1000, 1000, 1000, 
1000, 1000, 30, 30, 30, 30, 30, 30, 30, 30, 50, 50, 50, 50, 50, 
50, 50, 50, 75, 75, 75, 75, 75, 75, 75, 75, 100, 100, 100, 100, 
100, 100, 100, 100, 150, 150, 150, 150, 150, 150, 150, 150, 250, 
250, 250, 250, 250, 250, 250, 250, 500, 500, 500, 500, 500, 500, 
500, 500, 750, 750, 750, 750, 750, 750, 750, 750, 1000, 1000, 
1000, 1000, 1000, 1000, 1000, 1000, 30, 30, 30, 30, 30, 30, 30, 
30, 50, 50, 50, 50, 50, 50, 50, 50, 75, 75, 75, 75, 75, 75, 75, 
75, 100, 100, 100, 100, 100, 100, 100, 100, 150, 150, 150, 150, 
150, 150, 150, 150, 250, 250, 250, 250, 250, 250, 250, 250, 500, 
500, 500, 500, 500, 500, 500, 500, 750, 750, 750, 750, 750, 750, 
750, 750, 1000, 1000, 1000, 1000, 1000, 1000, 1000, 1000), mean_err = c(0.532157164015659, 
0.0344907054484091, 1.09430750651166, 0.847845162156675, 1.72337372126503, 
0.893756658507988, 0.0430309247027736, 0.785821459035346, 0.368063404388512, 
0.0256197760404459, 0.54480529648463, 0.54385597929651, 0.657325657699579, 
0.578611116865437, 0.0400491342691897, 0.515688005217413, 0.256713912589642, 
0.0444582875885996, 0.326975493112402, 0.371128780921122, 0.473515115741104, 
0.389519558405289, 0.105044360789378, 0.301924717299333, 0.207750921776918, 
-0.00318128936770314, 0.215170156089776, 0.27218780048926, 0.260519564021842, 
0.236792923882582, 0.0209319074923902, 0.232957055204834, 0.0726605917614469, 
-0.00282823355849125, 0.0753216783313991, 0.147121931849656, 
0.0986417724955371, 0.10048009778446, -0.0117861260923649, 0.0650845904350442, 
0.0985144485083747, 0.00601003227249322, 0.107439979908118, 0.120777421732797, 
0.098470489820427, 0.0940946984227826, 0.0340166854141625, 0.107851118082414, 
0.0249685210781582, -0.00275219600614378, 0.0258069390207201, 
0.0303381972274654, -0.000994602151869198, 0.00196854833683764, 
-0.0172319489562159, 0.0230957871932473, 0.00787486424680835, 
-0.018438041997315, -0.000808033567394628, -0.00151904153864496, 
-0.0258118523805697, -0.020984156892953, -0.0411584927473141, 
-0.0462075435919094, 0.0481149217843661, -0.0115241085997692, 
0.0278045708419731, 0.0358588316426625, 0.0424473909450939, 0.0379773233328197, 
0.0298772985879321, 0.0494939492036379, 0.561286668337778, 0.0210874502384648, 
0.607711822769155, 0.944733906256477, 1.32069801051061, 0.623604782428556, 
0.0138075109769806, 0.640859854412358, 0.284873604303057, -0.0170357985365701, 
0.273426417633118, 0.524432737895336, 0.355003110979807, 0.312837607951434, 
-0.0316296929553873, 0.270109834098986, 0.174110335581819, 0.0253101719615279, 
0.199956222742702, 0.375485416120771, 0.278956745944806, 0.244245525945888, 
0.0325126314233263, 0.253352659868514, 0.154840760004461, 0.00231589709639472, 
0.154412480165179, 0.264847742842386, 0.196572744185608, 0.182934783774992, 
0.0207139021497755, 0.17351041376412, 0.14190910348156, -0.000766834484010096, 
0.159975205214477, 0.191222128926019, 0.119768252112669, 0.12033372914036, 
-0.00429047209392067, 0.120982527821078, 0.0756304869484406, 
-0.00890884219048113, 0.0727782693168392, 0.118690444738942, 
0.0814898789647033, 0.0799724348001957, 0.0182926240912726, 0.0887155007804252, 
0.00799604720502299, 0.000599148388616836, -0.00567769035990384, 
0.0358412514670032, 0.0308474979074875, 0.0341668723768997, -0.000180318451026095, 
0.0180733341290925, 0.00456876236626807, 0.00150574966485876, 
0.011152095953916, 0.0176039119729626, 0.00608274255434991, 0.0146257828313115, 
-0.0108877417404102, 0.00341198000323035, -0.00198459880370283, 
0.0026551895445694, 0.00199040664534129, 0.0150165794544221, 
0.00646287144368147, 0.00999205240904708, -0.00850278571195971, 
0.000833666619266177, 0.714730067273087, -0.00916546079360956, 
0.662799490366986, 1.18552468844156, 1.25901933062308, 0.672701515532179, 
0.0311066140197676, 0.699068058809396, 0.451908043813962, 0.0131716226592205, 
0.429551369887738, 0.697928133235757, 0.445367768988423, 0.408463448982185, 
0.0318707241721211, 0.406284953982951, 0.257736307754364, -0.00525924423719458, 
0.236977000055322, 0.429144726596141, 0.291381752107184, 0.267557606428613, 
0.0103657879176852, 0.254728590646094, 0.187783398487578, -0.00216064381479362, 
0.209025103860707, 0.318293592390017, 0.216751610346408, 0.195630579126633, 
-0.00355723971644246, 0.174111826408428, 0.151010324964235, 0.0152409223899092, 
0.159002511320467, 0.214643583389694, 0.136211731513269, 0.138948149207635, 
0.00736196817594524, 0.115637867729083, 0.0491348055596302, -0.00133957946235943, 
0.0507437758212659, 0.103956325245849, 0.0641182216839426, 0.0721933081297794, 
-0.0124376134651938, 0.0632317888879588, 0.0322195712438111, 
0.00170122889182022, 0.0287526766624194, 0.0589662164030242, 
0.0348535721709848, 0.039527944642463, -0.00617539706415593, 
0.0274246010641889, 0.0325877909680824, 0.00530528253248245, 
0.0221776555499961, 0.0389702052631117, 0.0221602091288215, 0.0254478639695596, 
-0.0016189234058987, 0.0197144417326668, -0.00632485262604172, 
-0.00364979854195596, -0.00276076468388984, 0.0126267527301874, 
0.0123592498266038, 0.0154921632247644, -0.00591512196680815, 
0.0098685016547149, 1.19276916750486, -0.0296831640401583, 0.99406393593888, 
1.95758669445116, 1.45842790978446, 0.916899913902239, -0.0240222410217233, 
1.00771193034927, 0.748151091428865, 0.001671855025917, 0.665180535306263, 
1.1777049634557, 0.578603609273264, 0.546625362141714, 0.0292152981607387, 
0.615230814912951, 0.417886753756131, 0.00324593885807739, 0.406076310942717, 
0.732191741449251, 0.352684769616612, 0.326901376027897, -0.000759357576337989, 
0.350075431324921, 0.310927617707656, -0.0107255472998434, 0.28102101085112, 
0.514683023356017, 0.24913510139508, 0.235155452507568, -0.0220885572014814, 
0.243370611433649, 0.209652330609093, -0.00502865663759991, 0.2172246261346, 
0.356540958804122, 0.172121720418057, 0.17487914828986, 0.00365942442127361, 
0.176594681455494, 0.126956927057327, -0.00270525933073803, 0.120116234221594, 
0.210827536708082, 0.101520409193932, 0.101379097920023, 0.00238043252144371, 
0.113027315928011, 0.0598624378953727, -0.00538838415690431, 
0.0568400730102315, 0.0978115258288965, 0.0454207684906316, 0.0473140143579152, 
-0.00165813015281622, 0.0521772135812508, 0.0530224090961669, 
-0.000416993405198653, 0.0353236911458531, 0.0605493601241619, 
0.0316204159297213, 0.0344789374555544, -0.00446984887315054, 
0.0328807595695966, 0.0396438546423947, -0.00331466369719113, 
0.0379029847219126, 0.0572435100638761, 0.0253269328104989, 0.0235663211070417, 
0.00220241478536399, 0.0307132312422208), youden = c(0.2, 0.2, 
0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 
0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 
0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 
0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 
0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 0.2, 
0.2, 0.2, 0.2, 0.2, 0.2, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 
0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 
0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 
0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 
0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 
0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.4, 0.6, 
0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 
0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 
0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 
0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 
0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 
0.6, 0.6, 0.6, 0.6, 0.6, 0.6, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 
0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 
0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 
0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 
0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 
0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8, 0.8
)), row.names = c(NA, -288L), group_sizes = c(8L, 8L, 8L, 8L, 
8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 
8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L, 8L
), biggest_group_size = 8L, class = c("grouped_df", "tbl_df", 
"tbl", "data.frame"), groups = structure(list(sim_nr = 1:36, 
    .rows = list(1:8, 9:16, 17:24, 25:32, 33:40, 41:48, 49:56, 
        57:64, 65:72, 73:80, 81:88, 89:96, 97:104, 105:112, 113:120, 
        121:128, 129:136, 137:144, 145:152, 153:160, 161:168, 
        169:176, 177:184, 185:192, 193:200, 201:208, 209:216, 
        217:224, 225:232, 233:240, 241:248, 249:256, 257:264, 
        265:272, 273:280, 281:288)), row.names = c(NA, -36L), class = c("tbl_df", 
"tbl", "data.frame"), .drop = TRUE))
```

```{r, echo = FALSE}
library(ggplot2)
ggplot(plotdat_nomidpoints %>% filter(!(method %in% c("spline_20"))), 
       aes(x = n, y = mean_err, color = method, shape = method)) + 
    geom_line() + geom_point() +
    facet_wrap(~ youden, scales = "fixed") +
    scale_shape_manual(values = 1:nlevels(plotdat_nomidpoints$method)) +
    scale_x_log10(breaks = c(30, 50, 75, 100, 150, 250, 500, 750, 1000)) +
    ggtitle("Bias of all methods when use_midpoints = FALSE",
            "normally distributed data, 10000 repetitions of simulation")
```
